% !TEX TS-program = xelatex
% !BIB TS-program = bibtex
\documentclass[12pt,letterpaper]{article}
\usepackage{style/dsc180reportstyle} % import 
\usepackage{footmisc}
\usepackage{pdfpages}
% dsc180reportstyle.sty

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Title and Authors
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Hardware Acceleration for Tensor Multiplication in Neural Networks}

\author{Owen Shi \\
  {\tt a4shi@ucsd.edu} \\\And
  Pranav Kumarsubha \\
  {\tt pkumarsubha@ucsd.edu} \\\And
  Brendan Kuang \\
  {\tt blkuang@ucsd.edu} \\\And
  Edgar Guzman \\
  {\tt eiguzman@ucsd.edu} \\\And
  Idhant Kumar \\
  {\tt ikumar@ucsd.edu} \\\And
  Brandon Chiou \\
  {\tt 	brchiou@ucsd.edu} \\\And
  Rajesh Gupta \\
  {\tt rgupta@ucsd.edu} \\}

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Abstract and Links
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstract}
This checkpoint report describes our progress evaluating the linear-complexity multiplication algorithm (L-Mul) as an approximate alternative to standard floating-point multiplication for neural network inference. We implement L-Mul in Python, NumPy, and PyTorch and assess accuracy on multilayer perceptrons (MLPs), convolutional neural networks (CNNs), and long short-term memory networks (LSTMs); preliminary runs suggest L-Mul stays within 1\% of standard multiplication with high prediction agreement (e.g., 96.6\% in LSTM experiments). Ablations extend to nanoGPT, with perplexity on trained-from-scratch and pretrained models intended to characterize L-Mul's impact on autoregressive language modeling. At the system level, we integrate a custom L-Mul accelerator into the gem5 simulator and compare it to native IEEE BF16 matrix multiplication, with simulation time, cycle counts, and DRAM energy to be reported across matrix sizes and outputs validated against software references. We design and synthesize RTL for neural network linear layers in Vivado (with projected substantial area reduction versus IEEE multipliers while meeting timing) and target a tiled block matrix multiplication kernel for Qwen~2.5~0.5B on FPGA using BRAM-resident blocks to limit DRAM traffic. Projected outcomes, consistent with prior work, are that L-Mul preserves accuracy in software without surpassing optimized numerical libraries while hardware realizations yield meaningful efficiency gains; full results and discussion will follow in the final report.
\end{abstract}

\begin{center}
Code: \url{https://github.com/oowenn/LMUL-Hardware-Acceleration}
\end{center}

\maketoc
\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Main Contents
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

As artificial intelligence systems continue to scale in complexity and computational demand, speed and energy efficiency have emerged as critical constraints. Neural networks require billions of floating-point operations during inference, with traditional IEEE 754 multiplication incurring substantial computational overhead. The \textit{L-Mul} (Linear-complexity Multiplication) algorithm presents an alternative approach: replacing expensive mantissa multiplication with addition-based operations to achieve $O(m)$ time complexity compared to the $O(m^2)$ time complexity of IEEE.

This report presents our implementation and validation of the L-Mul algorithm for neural network inference. We focus specifically on BFloat16 (BF16) arithmetic, a 16-bit floating-point format widely adopted in machine learning due to its compatibility with FP32 exponents while requiring only half the storage. While algorithmic efficiency motivates L-Mul at the numerical level, its most compelling advantages emerge when considered in hardware. Conventional floating-point multipliers require complex mantissa multiplication units, normalization logic, and rounding circuitry, all of which contribute significantly to silicon area, power consumption, and design complexity. By replacing mantissa multiplication with addition-based operations, L-Mul enables a substantially simpler datapath that is better aligned with the resource and energy constraints of FPGA and accelerator-based systems. This architectural simplification creates an opportunity to co-design arithmetic, memory movement, and dataflow around neural network workloads, allowing us to explore how approximate multiplication can translate into system-level efficiency gains beyond software-only implementations.

On the software side, we investigate whether L-Mul is a practical substitute for standard floating-point multiplication in modern neural networks. We implement BF16 L-Mul in Python, NumPy, and PyTorch and compare results at both the scalar and tensor level, tracking approximation error and perplexity and analyzing how they affect model accuracy and stability during inference. By integrating L-Mul into linear and recurrent layers, we observe how these arithmetic changes propagate through full model computations. This is a checkpoint report; results and discussion will be presented in the final report. The remainder is organized as follows. Section~\ref{sec:background} reviews BF16 arithmetic, the L-Mul algorithm, and hardware tooling. Section~3 describes our methods: gem5 system simulation, Vivado RTL and synthesis, and ablations including nanoGPT. We summarize contributions and context at the end.

\section{Background}
\label{sec:background}

\subsection{Standard BFloat16 Multiplication}

\label{subsec:standard_bf16}

Before discussing the L-Mul approximation, we first establish how standard BFloat16 multiplication works. BFloat16 (BF16) is a 16-bit floating-point format consisting of 1 sign bit, 8 exponent bits, and 7 mantissa bits. This format is particularly popular in machine learning due to its compatibility with single-precision (FP32) exponents while requiring only half the storage.

A BF16 number $x$ represents the value:

\begin{equation}
x = (-1)^s \times 2^{e - 127} \times (1.m)
\label{eq:bf16_representation}
\end{equation}

where $s$ is the sign bit, $e$ is the 8-bit exponent, $m$ represents the 7-bit mantissa, and 127 is the bias.

Standard BF16 multiplication follows the IEEE 754 approach: the signs are XORed, the exponents are added (with bias correction), and the mantissas are multiplied. This mantissa multiplication is the expensive part—it requires a hardware multiplier and subsequent normalization logic to handle carries. For neural network inference where millions of multiplications occur, this computational cost adds up quickly in terms of both latency and power consumption.

\subsection{LMUL Algorithm}

\label{subsec:lmul_algorithm}

The L-Mul algorithm takes a different approach. Instead of treating the exponent and mantissa as separate entities, it exploits a clever observation: if we treat the combined exponent-mantissa field as a single fixed-point number, multiplication can be approximated by simple addition. This works because logarithms transform multiplication into addition, and the floating-point exponent already represents a logarithmic quantity.

\paragraph{Algorithm Overview.}

Given two BF16 operands $a$ and $b$, L-Mul works by adding their exponent-mantissa fields directly. Here's how it breaks down:

\textbf{Step 1: Extract the fields.} We pull out the components from each 16-bit operand:

\begin{align}
a_{sign} &= a[15], \quad b_{sign} = b[15] \\
a_{field} &= a[14:0], \quad b_{field} = b[14:0]
\end{align}

The $field$ values contain the concatenated exponent and mantissa bits, which we'll treat as a single 15-bit integer.

\textbf{Step 2: Handle edge cases.} If either input is zero or subnormal (indicated by a zero exponent), we immediately return zero. This keeps the algorithm simple and handles the most common edge case:

\begin{equation}
\text{if } a_{exp} = 0 \text{ or } b_{exp} = 0 \text{, return } 0
\end{equation}

\textbf{Step 3: Add the fields.} We add the two fields together, but we need to correct for the fact that adding exponents naturally doubles the bias:

\begin{equation}
\text{sum}_{full} = a_{field} + b_{field} + \text{OFFSET}_{MOD}
\label{eq:lmul_addition}
\end{equation}

The offset term compensates for this doubled bias:

\begin{equation}
\text{OFFSET}_{MOD} = ((2^{15}) - (127 \ll 7)) \text{ AND } 0x7FFF = 0x4080
\end{equation}

We use 17 bits for the sum to capture potential overflow.

\textbf{Step 4: Check for overflow or underflow.} The top two bits of our 17-bit sum tell us what happened:

\begin{equation}
\text{carry}_2 = \text{sum}_{full}[16:15]
\end{equation}

These bits give us three cases:

\begin{itemize}
    \item $00_2$: The result underflowed—return zero
    \item $01_2$: Normal result—use the lower 15 bits
    \item $1x_2$: The result overflowed—saturate to maximum value ($0x7FFF$)
\end{itemize}

\textbf{Step 5: Calculate the sign.} This follows standard multiplication rules—XOR the input signs, but force the result to positive zero if the magnitude is zero:

\begin{equation}
s_{result} = \begin{cases}
0 & \text{if field} = 0 \\
a_{sign} \oplus b_{sign} & \text{otherwise}
\end{cases}
\end{equation}

\textbf{Step 6: Pack the result.} Finally, we combine the sign bit with the selected field value:

\begin{equation}
\text{result} = (s_{result} \ll 15) \text{ OR } \text{field}_{sel}
\end{equation}

\paragraph{Why This Works.}

The key insight is that the exponent field already represents a logarithm (base 2), and the mantissa can be approximated as part of this logarithmic representation. When we add these combined fields, we're essentially performing multiplication in log space. The approximation comes from treating the mantissa bits as part of a continuous logarithmic scale rather than as a separate fractional multiplier. For neural network inference, this approximation can be accurate enough to preserve classification accuracy in practice.

\subsection{Multi Layer Perceptron}

To investigate the integration of approximate arithmetic operations within neural networks, we implement a modified Multi Layer Perceptron (MLP) in PyTorch. This MLP is designed to operate under both conventional floating-point modes and an approximate mode utilizing L-Mul, which models hardware-efficient multiplication. The primary goal was to directly replace the standard floating-point multiply operation in the network's forward pass with an approximate function that more accurately reflects the behavior of the custom hardware multiplier, thereby enabling an evaluation of the impact on model accuracy and inference speed.

\subsubsection{Model Architecture}

The architecture of the MLP used in our experiments comprises two fully connected (dense) layers interleaved with ReLU activation functions, culminating in a log-softmax output layer. Specifically, the model performs the following transformations:

\begin{itemize}
    \item \textbf{Input Layer}: Accepts a flattened 28 $\times$ 28 pixel grayscale image, resulting in a 784-dimensional input vector.
    \item \textbf{First Fully Connected Layer}: Maps the 784-dimensional input to a 128-dimensional hidden feature vector. This layer performs a matrix-vector multiplication with weight matrix $\mathbf{W}_1 \in \mathbb{R}^{128 \times 784}$ and bias vector $\mathbf{b}_1 \in \mathbb{R}^{128}$.
    \item \textbf{ReLU Activation}: Applies a rectified linear unit to introduce non-linearity.
    \item \textbf{Second Fully Connected Layer}: Projects the 128-dimensional hidden features onto a 10-dimensional output vector corresponding to class logits, via weight matrix $\mathbf{W}_2 \in \mathbb{R}^{10 \times 128}$ and bias vector $\mathbf{b}_2 \in \mathbb{R}^{10}$.
    \item \textbf{Log-Softmax Layer}: Converts the output logits into log-probabilities suitable for the negative log-likelihood loss.
\end{itemize}

\subsubsection{Mathematical Foundations}  
The core mathematical operations implemented in hardware replicate standard neural network computations. The linear layers perform dot products as follows:  
\begin{equation}
\mathbf{z} = \mathbf{W} \mathbf{x} + \mathbf{b}
\end{equation}  
where $\mathbf{W}$ is the weight matrix, $\mathbf{x}$ is the input vector, and $\mathbf{b}$ is the bias vector. The activation function used is ReLU, which is defined as:  
\begin{equation}
\text{ReLU}(x) = \max(0, x)
\end{equation}  
This ensures non-linearity and sparsity in activations before passing to subsequent layers.

\subsection{Hardware Integration}

In this subsection, we discuss the various hardware tools, simulation environments, and methodologies employed to develop, verify, and optimize our hardware acceleration modules for neural network computations, specifically focusing on tensor multiplication operations.

\subsubsection{gem5}
gem5 is a modular, open-source computer-architecture simulator widely used in research and industry to model full systems or syscall-emulation environments. It supports multiple ISAs (e.g., ARM, x86), detailed CPU and memory models, and custom devices, making it suitable for evaluating how an accelerator interacts with the rest of the system, including CPU, caches, and main memory, rather than in isolation. In our project we use gem5 to integrate a custom LMUL accelerator as a memory-mapped device, run the same BF16 matrix-multiplication workload on the accelerator and on the CPU (native IEEE), and compare simulation time, cycle counts, and DRAM energy; see Section~\ref{subsec:gem5} for the experimental setup and workflow.


\subsubsection{Icarus Verilog}

Icarus Verilog (iVerilog) is an open-source HDL simulation and synthesis tool that allows for the simulation of Verilog designs without requiring expensive proprietary software. It provides a lightweight environment suitable for rapid design iteration and testing \cite{iverilog}. The primary simulation tool associated with iVerilog is \texttt{vvp}, which executes compiled Verilog code (via \texttt{iverilog} compiler) and produces logs and debug information.

For our project, iVerilog and \texttt{vvp} served as accessible platforms for initial verification of test bench modules such as the multiplier (\texttt{lmul\_bf16.v}) and the Design Under Test (\texttt{top\_lmul.v}). Its python-like scripting capabilities make it suitable for engineers who are unfamiliar with high-level hardware design languages. iVerilog supports behavioral simulation, allowing us to test the correctness of our modules in a controlled environment before progressing to more advanced FPGA synthesis tools.

\subsubsection{Vivado}

AMD Xilinx Vivado Design Suite is an integrated development environment (IDE) tailored for FPGA synthesis, implementation, and debugging \cite{vivado}. Unlike iVerilog, which focuses on simulation, Vivado enables the compilation of HDL designs into FPGA bitstreams, providing a pathway from test bench simulation to RTL code to actual hardware deployment.

The advantages of Vivado in our work are manifold: it offers advanced synthesis optimizations, resource utilization analysis, and the ability to generate programming files for Xilinx FPGAs. Its graphical user interface facilitates design visualization and optimization through its analog waveform signal graphs. Employing Vivado allows us to validate both test bench and RTL modules within the context of actual FPGA hardware, enabling performance benchmarking, hardware debugging, and eventual deployment of the neural network acceleration modules.

\subsubsection{Test Bench Simulation}

Test Bench simulation is a fundamental step in verifying HDL modules, where a dedicated test environment (test bench) stimulates the RTL with input signals and monitors the outputs. Our test bench, \texttt{tb\_lmul.v}, simulates the behavior of the floating-point multiplier and the neural network accelerator modules.

The benefits of test bench simulation include the ability to perform comprehensive verification of logic correctness, timing analysis, and protocol adherence in a controlled setting. Additionally, simulation waveforms generated during test bench runs allow engineers to visualize signal transitions, identify glitches, and troubleshoot issues effectively. This process is essential for optimizing performance and ensuring the reliability of the hardware modules before hardware synthesis, minimizing costly iterations on physical FPGA devices.

\subsubsection{Register-Transfer Level (RTL)}

Register-Transfer Level (RTL) is an abstraction level used to describe digital circuits, focusing on the flow of data between registers and the logical operations performed on that data. RTL descriptions are typically written in hardware description languages such as Verilog or VHDL and serve as the primary input for synthesis tools.

In our project, RTL modules like \texttt{lmul\_bf16.v} and \texttt{fc\_accel.v} define the core computational units (multipliers, adders, and control logic) at a detailed hardware level. Using RTL allows us to specify precise hardware behavior, optimize resource utilization, and facilitate synthesis into FPGA fabric. The RTL approach ensures that the hardware implementation aligns with the intended data flow and control sequences necessary for efficient neural network acceleration.

\vspace{1em}
In summary, these tools enable us to validate correctness, optimize performance, and smoothly transition from HDL code to real hardware deployment, ultimately contributing to faster and more efficient neural network computations.

\section{Methods}

\subsection{Development Environment and Docker Configuration}

Docker was used for containerization to ensure consistent package reliability across different machines. Using the relevant files, a VS Code pipeline was developed to automate the installation of necessary Python, Verilog, and gem5 dependencies, while ensuring compatibility, as well as relevant VS Code extensions such as Python and Jupyter.

\subsection{Gem5 System Simulation}
\label{subsec:gem5}

To evaluate the practicality of an LMUL accelerator as a separate hardware component in a full-system context, we integrated a custom device model into the gem5 simulator and compared its performance to a native IEEE BF16 implementation on the same workload. The goal is to run identical BF16 matrix multiplications then compare simulation time, CPU cycles, and DRAM energy, and optionally validate outputs against software references.

Simulation mode, ISA, and workload (dense BF16 matrix multiply $C = A \times B$) are summarized in Tables~\ref{tab:gem5_config} and~\ref{tab:gem5_workflow}. The LMUL path programs the accelerator, streams $A$ and $B$ via memory-mapped I/O, polls for completion, and reads $C$ back. The IEEE path runs a naive scalar triple-loop on the CPU, performing each multiply in single precision by converting BF16 operands to float before multiplying and converting the result back to BF16. The same benchmark executable and dimensions are used for both runs, with identical initialization of $A$ and $B$, so results are comparable.

\paragraph{Data path: MMIO (LMUL) vs.\ main memory (IEEE).}
Currently our accelerator communicates using \emph{memory-mapped I/O} (MMIO): the device is exposed at fixed addresses, and the CPU streams $A$ and $B$ to the device and reads $C$ back by executing a load or store for each word, so the accelerator never touches main memory. To capture more realistic performance, a production design would use \emph{direct memory access} (DMA), where the CPU programs a transfer once and the accelerator or a DMA engine moves blocks between main memory and the device without the CPU touching every word. Meanwhile, the IEEE path uses no accelerator and only needs main memory: the CPU performs the matmul in software and accesses $A$, $B$, and $C$ via normal loads and stores through the memory system.

\begin{table}[H]
\centering
\caption{gem5 simulation and accelerator configuration.}
\label{tab:gem5_config}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Item} & \textbf{Setting} \\
\midrule
Simulation mode & Syscall emulation (SE), ARM \\
CPU & TimingSimpleCPU (single core) \\
Memory & Coherent bus, DDR3-style controller \\
Address space & 512\,MB \\
Accelerator interface & Memory-mapped I/O (MMIO) \\
PE array & Configurable (e.g.\ 4$\times$4, 8$\times$8) \\
LMUL data path & CPU streams $A$/$B$ in, reads $C$ via MMIO \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Benchmark and comparison workflow.}
\label{tab:gem5_workflow}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Item} & \textbf{Setting} \\
\midrule
Workload & Dense BF16 matrix multiply $C = A \times B$ \\
Matrix dimensions & Configurable (e.g.\ 4$\times$4 to 512$\times$512) \\
Runs per comparison & Two: LMUL (accelerator) and IEEE (CPU) \\
Performance metrics & Sim time, CPU cycles, instructions, CPI, IPC, DRAM energy \\
Accelerator stats & Total cycles, total ops, completions \\
Correctness (optional) & Compare outputs to \texttt{rtl.numpy\_lmul} (LMUL) or IEEE matmul (CPU) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Simulation and RTL Design in Vivado}

\subsubsection{Input Data Preparation}  
Our approach begins with generating the input data using a pretrained PyTorch classifier trained on the MNIST dataset. Specifically, we utilize a simple multi-layer perceptron (MLP) model with one hidden layer of 128 units. The trained model's weights and biases are extracted directly from PyTorch and formatted as 16-bit floating-point BF16 representations. These model parameters are then written into memory initialization files compatible with Vivado, residing in the designated simulation folders.  

Concurrently, MNIST test images and labels are processed using a Python script that converts each pixel value into BF16 format. The images, originally in 8-bit grayscale, are normalized to [0, 1], then converted into BF16 and stored sequentially in text files \texttt{test\_x.mem}. Labels are similarly stored in separate files \texttt{test\_y.mem}. These data files serve as inputs for both simulation and RTL implementation, ensuring consistency across the hardware and software components.

\subsubsection{Simulation Files}  
The simulation environment employs three key Verilog files:  
\begin{itemize}
    \item \textbf{\texttt{lmul\_bf16.v}}: Implements a pipelined BF16 multiplier. It accepts two BF16 inputs, performs the multiplication, and produces a BF16 output, handling handshakes for flow control.
    \item \textbf{\texttt{bf16\_adder.v}}: Performs addition of two BF16 operands, producing a BF16 result. This module models the core arithmetic operation of the dot product accumulations.
    \item \textbf{\texttt{tb\_lmul.v}}: A test bench that interfaces with the multiplier, feeding in input pairs, capturing output, and orchestrating the sequence of operations to simulate the neural network's linear layers. The logits generated are written to memory files for subsequent classification steps.
\end{itemize}  
These files collectively emulate the core computations of the neural network's fully connected layers in a behavioral simulation.

\subsubsection{Test Bench Integration}  
Implementing the test bench within Vivado offers significant advantages over traditional iVerilog simulations. Vivado facilitates timing analysis and resource utilization reports, essential for assessing the hardware acceleration potential. Viewing waveform graphs is crucial because it provides a detailed timeline of signal behavior, allowing verification that each clock cycle aligns with design specifications. It helps identify the precise moments when inputs are active or changing, ensuring correct sequencing and timing relationships. This visual insight is essential for diagnosing timing issues, bottlenecks, and confirming that the design operates as intended before implementation.

\subsubsection{RTL Implementation}  
To transition from behavioral simulation to hardware synthesis, we employ the \textbf{\texttt{fc\_accel.v}} module. This RTL block encapsulates the entire inference process, including storing input images, weights, biases, and performing the matrix-vector multiplications via instantiated multiplier and adder modules.  

The RTL design reorganizes the test bench into a synthesizable dataflow graph, implementing control logic to orchestrate data movement, multiplication, accumulation, and bias addition. This conversion allows us to synthesize the design in Vivado, enabling analysis of power consumption, processing speed, and resource utilization. The resultant logits are exported to software for final classification, maintaining the integrity of the inference pipeline. 

\subsection{Hardware Synthesis and System-Level Evaluation}

Building on the RTL design and simulation infrastructure described in previous sections, this work focuses on synthesizing both L-Mul and IEEE BF16 multipliers to obtain quantitative hardware metrics and extending the analysis to system-level accelerator configurations. The primary objectives are to measure area, timing, power, and throughput characteristics that validate L-Mul's efficiency advantages and establish a methodology for scaling from unit-level to accelerator-level evaluation.

\subsubsection{Synthesis Toolchain and Library}
Hardware synthesis is performed using \textbf{Yosys}, an open-source synthesis tool, in conjunction with the \textbf{Nangate 45nm Open Cell Library}. Yosys converts the RTL Verilog descriptions into gate-level netlists, mapping high-level operations to standard cells such as AND gates, OR gates, flip-flops, and adders. The synthesis process follows a structured flow:
\begin{enumerate}
    \item \textbf{RTL Elaboration}: Verilog modules (\texttt{lmul\_bf16.v} and \texttt{bf16\_mul.v}) are read and elaborated.
    \item \textbf{Optimization Passes}: Generic optimization passes (\texttt{proc}, \texttt{opt}, \texttt{fsm}, \texttt{memory}, \texttt{flatten}) are applied to simplify the design.
    \item \textbf{Technology Mapping}: The \texttt{abc} command maps combinational logic to standard cells, targeting a 1\,ns clock period (1\,GHz) to push timing limits.
    \item \textbf{Statistics Extraction}: Area and cell count statistics are extracted from the synthesis log using the \texttt{stat} command.
\end{enumerate}

Timing analysis is performed using \textbf{OpenSTA} (Open Static Timing Analyzer), which verifies that both designs meet the target clock constraint of 2.0\,ns (500\,MHz). Critical path delays, worst negative slack (WNS), and total negative slack (TNS) are extracted to ensure timing closure.

The L-Mul and IEEE BF16 multipliers are synthesized independently to establish baseline metrics. The synthesis scripts (\texttt{synth\_lmul.ys} and \texttt{synth\_bf16.ys}) apply identical optimization and mapping strategies to both designs. Key metrics extracted include:
\begin{itemize}
    \item \textbf{Chip Area}: Measured in library-specific units, representing the physical silicon footprint.
    \item \textbf{Cell Count}: Total number of standard cells instantiated, indicating design complexity.
    \item \textbf{Cell Breakdown}: Distribution of cell types (adders, logic gates, flip-flops, multipliers) to understand resource composition.
    \item \textbf{Timing Metrics}: Maximum clock frequency, critical path delay, and slack values.
\end{itemize}

\subsubsection{System-Level Design: PE Array}
To evaluate the scalability of L-Mul in a realistic accelerator context, we design a processing element (PE) array for both L-Mul and IEEE multipliers. The PE array is a 4$\times$4 systolic-style grid, where each PE performs a multiply-accumulate (MAC) operation. The Verilog module \texttt{pe\_array\_simple.v} instantiates 16 multipliers and 16 accumulators, along with control logic for data routing and synchronization.

The PE array design includes:
\begin{itemize}
    \item \textbf{Data Buffers}: Simple memory elements to store input matrices and partial results.
    \item \textbf{Control FSM}: A finite state machine that orchestrates the flow of data through the array, handles handshakes, and triggers accumulation.
    \item \textbf{Interconnect}: Routing logic to connect PEs and enable data reuse across columns and rows.
\end{itemize}

This subsystem-level design is synthesized using the same Yosys flow, and total area, cell count, and timing are measured. The goal is to understand how the unit-level area savings translate to system-level resource utilization and to estimate the number of PEs that can fit within a fixed silicon budget.

\subsubsection{Power Analysis Methodology}
While synthesis provides area and timing metrics, power consumption must be estimated through activity-aware analysis. Our power estimation methodology involves:
\begin{enumerate}
    \item \textbf{Workload Simulation}: The synthesized netlist is simulated using representative neural network workloads (e.g., MNIST inference, matrix multiplications) in a Verilog simulator (Icarus Verilog or Vivado).
    \item \textbf{VCD Generation}: During simulation, switching activity for all signals is recorded in a Value Change Dump (VCD) file. This captures the dynamic behavior of the circuit under real operating conditions.
    \item \textbf{Power Estimation}: The VCD file, along with the synthesized netlist and library power models, is analyzed using power estimation tools (e.g., Yosys power estimation, Synopsys PrimeTime PX, or open-source alternatives like \texttt{vcd2power}). Both dynamic power (due to switching activity) and leakage power (static) are estimated.
    \item \textbf{Comparison}: Identical workloads are used for both L-Mul and IEEE designs to ensure a fair comparison. Power per operation and total energy consumption are calculated.
\end{enumerate}

This approach provides a realistic estimate of power efficiency, accounting for the reduced switching activity expected from L-Mul's simpler logic.

\subsubsection{Throughput and Speedup Estimation}
System-level throughput is modeled as:
\[
\text{Throughput} = N_{\text{units}} \times f_{\text{clock}} \times \text{ops/cycle}
\]
where $N_{\text{units}}$ is the number of parallel compute units, $f_{\text{clock}}$ is the operating frequency, and ops/cycle depends on the pipeline depth and data reuse patterns.

By leveraging L-Mul's area advantage, approximately \textbf{3$\times$ more L-Mul units} can be instantiated within the same chip area as IEEE multipliers. This enables higher parallelism and potentially \textbf{3$\times$ system-level speedup}, assuming compute-bound workloads and sufficient memory bandwidth. The actual speedup depends on Amdahl's Law and the fraction of computation that can be accelerated.

\subsubsection{Memory and Control Overhead}
Real accelerators require memory hierarchies, control logic, and I/O interfaces. To account for these overheads, we synthesize the PE array with simple memory buffers and control FSMs, and measure their contribution to total area and power. This provides a more realistic estimate of system-level resource utilization. For full-system projections, we extrapolate from the PE array results, assuming proportional scaling of memory and control logic with the number of compute units.


\subsection{Tiled Block Matric Multiplication with Qwen 2.5-0.5B}
The primary target for computation will be a Tiled Block Matrix Multiplication kernel. Because LLM weight matrices are too large to fit entirely in the FPGA's BRAM, we partition the matrices into smaller $B \times B$ blocks. This strategy allows the hardware to load a block of weights once and reuse it across multiple vector operations, limiting the frequency off chip DRAM accesses. We will implement two versions of this kernel: a version using  IEEE-754 multipliers and a LMUL version.  By comparing these two versions, we aim to demonstrate that the LMUL can maintain the high-dimensional representational requirements of the Qwen-2.5 model while providing a more efficient hardware than traditional floating-point units.

\paragraph{PYNQ Integration and Development}
Consistent with our preliminary design, the Qwen accelerator operates on BF16. We maintain the model's original precision without quantization, because our LMUL operates on floating points. For software to communicate to the hardware, we will use PYNQ in Python. A Python script will use direct memory access to stream the Qwen weights to the FPGA's memory. 

\subsection{NanoGPT}

To further investigate L-Mul's impact on neural networks, we use NanoGPT, a minimal implementation of a GPT-2 style autoregressive transformer language model developed by Andrej Karpathy. This model was selected because of its reproducible GPT-style architecture with enough customizability to fit our L-Mul tensor operations into for experimentation. Its simplicity also enables us to isolate the effects of our L-Mul operation replacements for ablation testing. 

\paragraph{Model Architecture}
NanoGPT consists of: 
\begin{itemize}
    \item A token embedding layer, which takes tokens in their integer representation and outputs a float vector in a specified embedding dimension, also presented as the model's hidden width. 
    \item A stack of Transformer blocks combining masked multi-head self-attention and feed-forward (MLP) sublayers. 
    \item A linear langue modeling head that maps hidden states to vocabulary logits.
\end{itemize}
Our configuration of NanoGPT features the following parameters: a block size of 32, a vocabulary size of 50,304, 4 Transformer block layers, 4 self-attention heads per Transformer block, and an embedding dimension of 48. 
\paragraph{L-Mul Integration}
To utilize L-Mul within NanoGPT's model architecture, toggleable L-Mul Linear Layers were implemented in place of PyTorch's nn.Linear layer for forward passes. These L-Mul Linear Layers utilize the PyTorch L-Mul implementation listed in LSTM–L-Mul Integration. The three main architectures in which we can toggle L-Mul operations for are the language model head, Tranformer block feed-forward, and self-attention layers.
\paragraph{Shakespeare-Char Model (Controlled Training)}
To obtain a fully controlled training baseline, we train a NanoGPT language model from scratch on the Shakespeare-Char dataset. This dataset consists of character-level text sequences, enabling efficient training and rapid iteration on architectural modifications. 
\paragraph{OpenWebText Model (Pretrained Evaluation)}
To evaluate L-Mul within a more realistic large-corpus language modeling setting, we also use a pretrained nanoGPT model trained on OpenWebText. This pretrained checkpoint provides a much broader linguistic distribution and vocabulary compared to Shakespeare-Char, enabling for the evaluation of L-Mul under conditions more representative of general-purpose language modeling. The pretrained model is used strictly for inference and perplexity evaluation to assess how LMUL modifications affect an already-trained Transformer.
\paragraph{Evaluation}
Both L-Mul and base Model performances are evaluated using token-level cross-entropy and perplexity on held-out validation data. For the Shakespeare-Char dataset, validation batches are constructed by sampling random subsequences from the validation corpus. Because character-level modeling exhibits limited long-range dependency structure, random subsequence sampling provides a representative estimate of predictive performance while reducing evaluation bias.

For the OpenWebText model, validation is performed using sequences with fixed starting positions drawn sequentially from the validation corpus. This approach preserves the context required for stable perplexity estimation in word-level language modeling, where predictive accuracy depends strongly on long-range dependencies across tokens.

Controlled ablation experiments are performed by selectively enabling LMUL operations in attention projections, MLP layers, and output heads, allowing isolation of their contribution to predictive performance.

\subsection{Neural Network Ablation Studies}

Ablation studies were conducted to identify which components of each architecture are most sensitive to L-Mul substitution, rather than evaluating the full model in a single setting. Isolating sensitivity by layer or module informs hardware prioritization: components that tolerate approximation with minimal accuracy loss are natural candidates for dedicated L-Mul units.

\paragraph{LSTM Ablation.} The LSTM was divided into two logical parts: (1) the gate computation, i.e., the matrix multiplication that produces the input, forget, cell-candidate, and output gates ($i$, $f$, $g$, $o$) from the weight matrix and concatenated input; and (2) the state update, which updates the cell state and hidden state. A modified LSTM class was implemented with two boolean flags to enable or disable L-Mul independently in each part. Perplexity was compared on the Shakespeare dataset using pretrained weights. Preliminary results indicate that the state-update path performs slightly better with L-Mul enabled, while the gate path remains roughly unchanged.

\paragraph{CNN Ablation.} A similar decomposition was applied to the CNN, separating convolutional layers from fully connected (FC) layers, which perform distinct operations. The convolutional layers use \texttt{lmul\_conv2d}, which unfolds the input into patches and performs the convolution via L-Mul; the FC layers use \texttt{lmul\_linear}. In preliminary runs, the FC layers retained accuracy well under L-Mul, whereas the convolutional layers showed greater sensitivity. This is consistent with the fact that convolutions operate over the full spatial extent of the feature maps, so approximation errors can accumulate across positions. The findings suggest that hardware designs may prioritize L-Mul in dense (linear) layers before extending it to convolutional layers.

\paragraph{Website and Documentation.} A project website was developed to centralize algorithm descriptions (including the code used for L-Mul), tables of accuracy and hardware metrics (e.g., area reduction and cell count), and team contributions. A bar chart was added to improve readability of the accuracy comparisons. The site is updated as additional ablation and hardware evaluation results become available.

% \section{Results}

% \subsection{Gem5 Simulation}

% Summary statistics from the simulations are presented in Table~\ref{tab:gem5_results}.

% \begin{table}[H]
% \centering
% \caption{Summary of gem5 comparison across matrix sizes $n \in \{4, 64, 512\}$: LMUL accelerator (MMIO) vs.\ IEEE CPU-only BF16 matmul.}
% \label{tab:gem5_results}
% \begin{tabular}{@{}lrrrrrr@{}}
% \toprule
% & \multicolumn{2}{c}{$n = 4$} & \multicolumn{2}{c}{$n = 64$} & \multicolumn{2}{c}{$n = 512$} \\
% \cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7}
% \textbf{Metric} & LMUL & IEEE & LMUL & IEEE & LMUL & IEEE \\
% \midrule
% Simulation time (s) & 0.00055 & 0.00061 & 0.0152 & 0.202 & 0.930 & 95.8 \\
% CPU cycles & 1.09M & 1.22M & 30.3M & 404M & 1.86B & 191.5B \\
% Instructions & 8.2k & 9.1k & 199k & 2.82M & 12.2M & 1.35B \\
% CPI & 134.1 & 133.8 & 152.1 & 143.2 & 152.1 & 141.4 \\
% IPC & 0.0075 & 0.0075 & 0.0066 & 0.0070 & 0.0066 & 0.0071 \\
% DRAM energy ($\mu$J) & 667k & 743k & 17.4M & 239M & 1.07e9 & 1.17e11 \\
% \midrule
% \multicolumn{7}{l}{\textit{LMUL run only:}} \\
% Accelerator total cycles & 4 & --- & 16,460 & --- & 2.10M & --- \\
% Accelerator total ops & 64 & --- & 262k & --- & 134M & --- \\
% \bottomrule
% \end{tabular}
% \end{table}

% \section{Discussion}

% \subsection{Notes on Gem5 Simulation}

% The current gem5 setup lets us compare an LMUL accelerator (MMIO data path) against a CPU-only IEEE BF16 matmul on the same workload and dimensions. Any reported speedups should be interpreted as accelerator (MMIO) vs.\ naive CPU (scalar, no SIMD or tiling), not as a comparison with a DMA-based accelerator or an optimized CPU baseline. More practical designs would use DMA for bulk data transfer, which would change both absolute performance and the balance between transfer and compute. The IEEE path is intentionally naive (BF16$\to$float$\to$BF16 in a scalar triple-loop) to keep the comparison controlled; production CPU code would use SIMD, tiling, or dedicated BF16 support where available.

% The simulation can report \emph{DRAM energy} from the memory system, which reflects energy consumed by main memory when data passes through it. For the IEEE run, all matmul data ($A$, $B$, $C$) goes through the memory system, so DRAM energy is relevant. However, for the LMUL run, matmul data is streamed via MMIO and does not go through DRAM, so DRAM energy for that run is largely from other traffic. Since the accelerator device itself has no power model yet, we can not yet report on its energy or power.

% \subsection{Future Work}

% \subsubsection{Gem5 Simulation}
% In order to provide the most realistic evaluation of an LMUL accelerator, we plan to (i) move bulk data transfer off the CPU by implementing DMA so the accelerator or a DMA engine moves $A$, $B$, and $C$ between main memory and the device; (ii) strengthen the IEEE baseline by optimizing the CPU matmul path (e.g., native or emulated BF16 where possible, better loop structure or tiling) so speedups are not overstated against an intentionally weak baseline; (iii) categorize where cycles are spent (memory traffic, compute, MMIO, polling) in both runs so we can see how DMA and a stronger CPU baseline change the tradeoff; (iv) add a power model to the accelerator so we can compare energy cost in addition to latency; and (v) potentially incorporate silicon area evaluation, but this is not natively supported in gem5 and would have to be sourced from a separate synthesis.

\section{Contributions}
\begin{itemize}
    \renewcommand{\labelitemi}{--}
    \item Owen: Docker setup, lmul verilog files, gem5 simulation.
    \item Edgar: iVerilog tests, Vivado testbench simulation/optimization and RTL implementation.
    \item Pranav: Website development \& delivery, ablation testing \& optimization, LSTM \& CNN evaluation 
    \item Idhant: Hardware synthesis workflow (Yosys, OpenSTA), unit and system-level evaluation methodology, power analysis framework (VCD generation and estimation).
    \item Brendan: NanoGPT Training, ablation testing, and perplexity analysis. LSTM and MLP implementation. L-Mul Linear Layer implementation.
    \item Brandon: Tiled matrix multiplication in Vitis HLS,  LMUL algorithm into C++, qwen model loading.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
\makeappendix

\subsection{Context and Motivation}

As large language models (LLMs) and other large neural networks integrate into modern society, their applications are increasingly bottlenecked by computational intensity. The core operation driving these models is floating-point multiplication in matrices, which accounts for approximately 65--85\% of operations in LLMs like ChatGPT \citep{dettmers2022llmint8}. Given the scale at which these models operate, even slight improvements in each multiplication operation could yield substantial impacts on overall efficiency and performance. As model sizes continue to grow and deployment expands to energy-constrained edge devices, developing more efficient alternatives to traditional floating-point multiplication becomes critical for sustainable AI deployment.

\subsection{Problem Statement}

Enhancing floating-point multiplication has emerged as an active area of research, with several approaches exploring approximate arithmetic to reduce computational cost. Linear-complexity Multiplication (L-Mul) is an algorithm that estimates floating-point multiplication in $\mathcal{O}(m)$ time with respect to the size of the data type (mantissa bits), as opposed to the $\mathcal{O}(m^2)$ complexity of traditional multiplication \citep{luo2024addition}. This theoretical advantage suggests significant potential for speedup, energy savings and silicon area reduction in hardware implementations.

Previous work has validated performance and energy trade-offs of L-Mul in baseline tasks, particularly multilayer perceptrons (MLPs) trained on MNIST \citep{lmul2023capstone}. These results demonstrate that L-Mul can maintain accuracy within 1\% of standard floating-point multiplication, reduce area by 76.1\%, and reduce power by 96.6\%. However, this evaluation lacks generalizability to real-world applications, which are at a much larger scale in terms of size and complexity. The algorithm remains largely theoretical, having been proposed only recently, and has not yet seen comprehensive testing in practice across diverse neural network architectures and complex tasks. For our future evaluations, we will focus on FashionMNIST as our primary dataset for classification tasks. FashionMNIST provides 70,000 labeled images across 10 classes of clothing items, offering a more challenging classification problem than MNIST while maintaining similar structure and scale, making it suitable for evaluating L-Mul's effectiveness in practical image classification scenarios.

With L-Mul being proposed just over a year ago, the gap between theoretical promise and practical deployment may stem from several factors: the cost of implementing specialized hardware, challenges in operating system integration to achieve real performance metrics, and insufficient validation across the breadth of neural network architectures used in production systems.

\subsection{Primary Output}

Our project will produce a comprehensive analysis of L-Mul as a hardware accelerator for floating-point multiplication in deep neural networks. The primary deliverables include:

\begin{itemize}
    \item \textbf{Performance Analysis}: Quantitative evaluation of estimated changes in performance (speed, power, and accuracy) when accelerating key operations such as fp matrix multiplication in large neural networks using L-Mul, compared against standard floating-point multiplication implementations.
    
    \item \textbf{Accuracy Experimentation}: Validation of accuracy performance across complex tasks and architectures, including Transformers and LSTMs on FashionMNIST classification and sequential token generation tasks.
    
    \item \textbf{Cost and Obstacle Analysis}: Comprehensive evaluation of implementation costs and deployment challenges, including chip manufacturing costs, operating system compatibility requirements, hardware-software interface design, and integration complexity with existing neural network frameworks.
\end{itemize}
All analyses and experimental results will be communicated through a detailed technical report documenting methodology, findings, and recommendations for future work. Code implementations, hardware designs, and experimental notebooks will be made publicly available to facilitate reproducibility and further research.

\subsection{Path to Success}

Given our work in Quarter 1, we can ensure that data related to L-Mul benchmarks can be realistically obtained. We can run our experiments on local desktop computers equipped with sufficient GPU and memory resources, as well as on cloud platforms such as DSMLP. Preliminary tests confirm that data loading, preprocessing, and model training can be conducted efficiently, and that existing neural network implementations can be modified to incorporate L-Mul in both software and hardware settings. For hardware evaluation, Icarus Verilog (iverilog) provides simulation runtime and clock-cycle estimates, and Yosys with OpenSTA have been explored to yield cell area, cell count, timing metrics, and power estimates. Preliminary results from our MLP, LSTM, and CNN classification tests show promise for extending L-Mul to Transformers or LSTMs on classification as well as new tasks such as token generation or regression, with accuracy maintained within 1\% of standard floating-point implementations across diverse architectures in early runs.

\bibliography{reference}
\bibliographystyle{style/dsc180bibstyle}

\end{document}